Do we need detailed demographic data to forecast plant population responses to climate change?

### 2015-06-16 manuscript/manuscript.Rmd
adds results section; completed draft

In this paper, we model populations based on two levels of data: individual and quadrat (Figure 1). The individual data is the "raw" data. For the quadrat level we data we simply sum individual areal cover for each quadrat by species. This is equivalent to a perfect census of quadrat percent cover, so we do not need to consider measurement error. Based on these two datasets we can compare population models built using individual level data and aggregated quadrat level data.

@@

### Sensitivity of models to climate

@@

```{r figure_2, dependson="plot-options", fig.cap="Sensitivity of equilibrium cover to a 1% increase in precipitation (+ppt), temperature (+temp), or both (+ppt&temp) applied to individual and combined vital rates. For example, the points associated with G show the median cover from IPM simulations where a climate perturbation is applied only to the growth regression climate covariates.", fig.width=4.5, fig.height=6, cache=FALSE, results='hide'}
setwd("../analysis/ipm/simulations/results/climate_sensitivity/")
files <- list.files()
cover_files <- files[grep("cover", files)]
spp_id <- substr(cover_files, 1, 4)
num_files <- length(cover_files)
all_sims <- data.frame(time=NA, cover=NA, species=NA, climsim=NA, vital=NA)
for(i in 1:num_files){
  tmp <- read.csv(cover_files[i])
  tmp$species <- rep(spp_id[i], nrow(tmp))
  all_sims <- rbind(all_sims, tmp)
}
all_sims <- all_sims[2:nrow(all_sims),]

```{r figure_3, dependson="plot-options", fig.cap="Proportional change in species' mean cover caused by a 1% increase in observed precipitation (+ppt), temperature (+temp), or both (+ppt&temp) as predicted by the individual-based IPM and the aggregate-based QBM. Top panels show the mean predicted proportional change in cover; lower panels show the 90% quantiles of predicted proportional cover change. Mean predictions and their associated uncertainties are shown in separate panels for visual clarity.", fig.width=8.5, fig.height=6, cache=FALSE, results='hide'}
setwd("~/Repos/MicroMesoForecast/analysis/")

@@

>>>>

In this paper, we model populations based on two levels of data: individual and quadrat (Fig. 1). The individual data is the "raw" data. For the quadrat level we data we simply sum individual areal cover for each quadrat by species. This is equivalent to a perfect census of quadrat percent cover, so we do not need to consider measurement error. Based on these two datasets we can compare population models built using individual level data and aggregated quadrat level data.

@@

### Sensitivity of models to climate
```{r get_results, echo=FALSE, eval=TRUE}
setwd("~/Repos/MicroMesoForecast/analysis/")

##  Merge IPM and QBM mean parameter results
bothmodel_outs <- rbind(ipm_meansims, qbm_meansims)
meanparam_agg <- ddply(bothmodel_outs, .(species, climsim, model), summarise,
                       meandiff = mean(propdiff),
                       hidiff = quantile(propdiff, probs = 0.95),
                       lodiff = quantile(propdiff, probs = 0.05))
setwd("~/Repos/MicroMesoForecast/manuscript")
```

Equilibrium cover from both models was sensitive to climate (Fig. 2a-d). The IPM projected percent changes in equilibrium cover from `r round(min(subset(meanparam_agg, species=="BOGR" & model=="IPM")["meandiff"])*100)` - `r round(max(subset(meanparam_agg, species=="BOGR" & model=="IPM")["meandiff"])*100)`% for *B. gracilis*, `r round(min(subset(meanparam_agg, species=="HECO" & model=="IPM")["meandiff"])*100)` - `r round(max(subset(meanparam_agg, species=="HECO" & model=="IPM")["meandiff"])*100)`% for *H. comata*, `r round(min(subset(meanparam_agg, species=="PASM" & model=="IPM")["meandiff"])*100)` - `r round(max(subset(meanparam_agg, species=="PASM" & model=="IPM")["meandiff"])*100)`% for *P. smithii*, and `r round(min(subset(meanparam_agg, species=="POSE" & model=="IPM")["meandiff"])*100)` - `r round(max(subset(meanparam_agg, species=="POSE" & model=="IPM")["meandiff"])*100)`% for *P. secunda*. The QBM projected opposite and greater percent changes in equilibrium cover for *B. gracilis* (`r round(min(subset(meanparam_agg, species=="BOGR" & model=="QBM")["meandiff"])*100)` - `r round(max(subset(meanparam_agg, species=="BOGR" & model=="QBM")["meandiff"])*100)`%) and *H. comata* (`r round(min(subset(meanparam_agg, species=="HECO" & model=="QBM")["meandiff"])*100)` - `r round(max(subset(meanparam_agg, species=="HECO" & model=="QBM")["meandiff"])*100)`%; Fig. 2a-b). For *P. smithii*, the QBM projected opposite changes in equilibrium cover than the IPM, but of similar magnitude (`r round(min(subset(meanparam_agg, species=="PASM" & model=="QBM")["meandiff"])*100)` - `r round(max(subset(meanparam_agg, species=="PASM" & model=="QBM")["meandiff"])*100)`%; Fig. 2c). *P. secunda* was the only species that the IPM and QBM made projections of the same sign and somewhat similar magnitude (Fig. 2d). As expected based on model validation (Table 1), IPM projections were more uncertain than QBM projections for all species and all climate change scenarios (Fig. 2e-h).

The response of a population to climate change is a result of the aggregate effects of climate on individual vital rates. Since the IPM approach relies on vital rate regressions, we were able to quantify the sensitivity of each vital rate in isolation and in pairwise combinations. Species showed similar trends (Fig. 3). Growth was the most sensitive vital rate for all species, showing a negative response to increased precipitation, and stronger positive response to increased temperature, and a mostly positive response when both climate factors are increased (Fig. 3). *B. gracilis* survival rates were sensitive to temperature, resulting in an increase in plant cover under increased temperature (Fig. 3a). In isolation, recruitment and survival were insensitive to climate factors for *H. comata* (Fig. 3b). Survival and recruitment of *P. smithii* were both sensitive, negatively, to temperature and precipitation (Fig. 3c). *P. secunda* equilibrium cover was sensitive to the climate effects on survival and recruitment, showing a negative effect on both vital rates for increased precipition, but a strong positive effect on survival with increased temperature (Fig. 3d). The climate impact of recruitment on equilibrium cover was negative for precipitation and temperature increases (Fig. 3d).

@@

```{r figure_2, dependson="plot-options", fig.cap="Proportional change in species' mean cover caused by a 1% increase in observed precipitation (+ppt), temperature (+temp), or both (+ppt&temp) as predicted by the individual-based IPM and the aggregate-based QBM. Top panels show the mean predicted proportional change in cover; lower panels show the 90% quantiles of predicted proportional cover change. Mean predictions and their associated uncertainties are shown in separate panels for visual clarity.", fig.width=8.5, fig.height=6, cache=FALSE, results='hide'}
setwd("~/Repos/MicroMesoForecast/analysis/")

@@

```{r figure_3, dependson="plot-options", fig.cap="Sensitivity of equilibrium cover to a 1% increase in precipitation (+ppt), temperature (+temp), or both (+ppt&temp) applied to individual and combined vital rates. For example, the points associated with G show the median cover from IPM simulations where a climate perturbation is applied only to the growth regression climate covariates.", fig.width=4.5, fig.height=6, cache=FALSE, results='hide'}
setwd("../analysis/ipm/simulations/results/climate_sensitivity/")
files <- list.files()
cover_files <- files[grep("cover", files)]
spp_id <- substr(cover_files, 1, 4)
num_files <- length(cover_files)
all_sims <- data.frame(time=NA, cover=NA, species=NA, climsim=NA, vital=NA)
for(i in 1:num_files){
  tmp <- read.csv(cover_files[i])
  tmp$species <- rep(spp_id[i], nrow(tmp))
  all_sims <- rbind(all_sims, tmp)
}
all_sims <- all_sims[2:nrow(all_sims),]

### 2015-06-16 manuscript/manuscript.tex
adds results section; completed draft

In this paper, we model populations based on two levels of data:
individual and quadrat (Figure 1). The individual data is the ``raw''
data. For the quadrat level we data we simply sum individual areal cover
for each quadrat by species. This is equivalent to a perfect census of
quadrat percent cover, so we do not need to consider measurement error.
Based on these two datasets we can compare population models built using
individual level data and aggregated quadrat level data.

@@

@@

We sought to test the assumption that the sensitivities of plant
populations to climate variables can be detected equally well using
either individual level data or population level data. This is an
important question to answer because population models are key tools for
predicting the consequences of global climate change. However, they can
be of limited use when built on data from a small subset of a population
in space or time. If population level data (i.e., some aggregated form
of individual level data) can be used to detect climate effects on
population dynamics, then we would have a cheaper and easier option for
data collection over relatively large temporal and spatial extents
(\emph{e.g.} Freckleton et al. 2011).

Our analysis offers good news and bad news. The good news is that for
two out of our four focal species, the QBM faired no worse than the
demographically detailed IPM in terms of mean absolute (Table 1). Thus,
it seems that aggregate level data can provide forecasts of similar
skill as individual level data.

@@

\begin{table}[ht]
\centering
\caption{Accuracy (mean absoluet error, MAE) and precision (90\% Distance) of out of sample predictions. Forecasts were made without random year effects; only climate covariates could explain year-to-year variation. 90\% Distance refers to the average distance between the upper and lower 90th percentiles of the 100 predicted values for each quadrat-year combination.} 
\begin{tabular}{llrrr}
  \hline
Species & Model & MAE & 90\% Distance & Mean Obs. Cover \\ 
  \hline
BOGR & IPM & 12.18 & 38.52 & 9.43 \\ 
  BOGR & QBM & 19.66 & 56.50 & 9.26 \\ 
  HECO & IPM & 1.22 & 6.47 & 1.15 \\ 
  HECO & QBM & 12.35 & 41.11 & 1.18 \\ 
  PASM & IPM & 0.19 & 1.65 & 0.42 \\ 
  PASM & QBM & 0.55 & 7.78 & 0.42 \\ 
  POSE & IPM & 1.37 & 7.64 & 1.25 \\ 
  POSE & QBM & 1.79 & 40.59 & 1.27 \\ 
   \hline
\end{tabular}
\end{table}

@@

\begin{figure}[htbp]
\centering
\includegraphics{components/figure/manuscript-figure_2.pdf}
\caption{Sensitivity of equilibrium cover to a 1\% increase in
precipitation (+ppt), temperature (+temp), or both (+ppt\&temp) applied
to individual and combined vital rates. For example, the points
associated with G show the median cover from IPM simulations where a
climate perturbation is applied only to the growth regression climate
covariates.}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics{components/figure/manuscript-figure_3.pdf}
\caption{Proportional change in species' mean cover caused by a 1\%
increase in observed precipitation (+ppt), temperature (+temp), or both
(+ppt\&temp) as predicted by the individual-based IPM and the
aggregate-based QBM. Top panels show the mean predicted proportional
change in cover; lower panels show the 90\% quantiles of predicted
proportional cover change. Mean predictions and their associated
uncertainties are shown in separate panels for visual clarity.}
\end{figure}

@@

Clark, J. S., D. M. Bell, M. Kwit, A. Stine, B. Vierra, and K. Zhu.
2012. Individual-scale inference to anticipate climate-change
vulnerability of biodiversity.

>>>>

In this paper, we model populations based on two levels of data:
individual and quadrat (Fig. 1). The individual data is the ``raw''
data. For the quadrat level we data we simply sum individual areal cover
for each quadrat by species. This is equivalent to a perfect census of
quadrat percent cover, so we do not need to consider measurement error.
Based on these two datasets we can compare population models built using
individual level data and aggregated quadrat level data.

@@

Equilibrium cover from both models was sensitive to climate (Fig. 2a-d).
The IPM projected percent changes in equilibrium cover from -3 - 8\% for
\emph{B. gracilis}, -4 - 3\% for \emph{H. comata}, -15 - 9\% for
\emph{P. smithii}, and -17 - 53\% for \emph{P. secunda}. The QBM
projected opposite and greater percent changes in equilibrium cover for
\emph{B. gracilis} (-63 - 30\%) and \emph{H. comata} (-50 - -18\%; Fig.
2a-b). For \emph{P. smithii}, the QBM projected opposite changes in
equilibrium cover than the IPM, but of similar magnitude (-5 - 6\%; Fig.
2C). \emph{P. secunda} was the only species that the IPM and QBM made
projections of the same sign and somewhat similar magnitude (Fig. 2d).
As expected based on model validation (Table 1), IPM projections were
more uncertain than QBM projections for all species and all climate
change scenarios (Fig. 2e-h).

The response of a population to climate change is a result of the
aggregate effects of climate on individual vital rates. Since the IPM
approach relies on vital rate regressions, we were able to quantify the
sensitivity of each vital rate in isolation and in pairwise
combinations. Species showed similar trends (Fig. 3). Growth was the
most sensitive vital rate for all species, showing a negative response
to increased precipitation, and stronger positive response to increased
temperature, and a mostly positive response when both climate factors
are increased (Fig. 3). \emph{B. gracilis} survival rates were sensitive
to temperature, resulting in an increase in plant cover under increased
temperature (Fig. 3a). In isolation, recruitment and survival were
insensitive to climate factors for \emph{H. comata} (Fig. 3b). Survival
and recruitment of \emph{P. smithii} were both sensitive, negatively, to
temperature and precipitation (Fig. 3c). \emph{P. secunda} equilibrium
cover was sensitive to the climate effects on survival and recruitment,
showing a negative effect on both vital rates for increased precipition,
but a strong positive effect on survival with increased temperature
(Fig. 3d). The climate impact of recruitment on equilibrium cover was
negative for precipitation and temperature increases (Fig. 3d).

@@

Demographic models are costly to parameterize, yet remain the most
widely-used tools for studying plant population dynamics (e.g.,
Salguero-Gómez et al. (2015)). New density-structured approaches
sacrifice some precision at the local scale for better spatial coverage
that is virtually unattainable, due to time and funding constraints,
using traditional individual-based data (Freckleton et al. 2011,
Queenborough et al. 2011). These approaches are appealing because
traditional population models suffer from constrained parameter space --
meaning forecasts quickly extend beyond the parameter space used to fit
the model. Density-structured approaches rely on easy-to-collect census
data (e.g., percent cover in $1\text{m}^2$ quadrats) so sampling
populations across large spatial extents would be possible, thus
covering more potential parameter space.

Estimating parameter variability in space and time is critical when
using population models to forecast responses to directional changes in
exogoneous drivers like climate ({\textbf{???}}, {\textbf{???}},
{\textbf{???}}, {\textbf{???}}). Compared to demographic models,
density-structured modeling approaches may provide an efficient route
toward estimating such uncertainty. However, population models based on
aggregated inidividual data have yet to be tested with climate drivers.
Since it is individuals that respond to weather, rather than
populations, models based on a aggregated data may produce incorrect
forecasts (Clark et al. 2012). Thus, we sought to compare climate
sensitivities of individual and population based models of four
grassland species populations.

@@

\begin{table}[ht]
\centering
\caption{Accuracy (mean absolute error, MAE) and precision (90\% Distance) of out of sample predictions. Forecasts were made without random year effects; only climate covariates could explain year-to-year variation. 90\% Distance refers to the average distance between the upper and lower 90th percentiles of the 100 predicted values for each quadrat-year combination.} 
\begin{tabular}{llrrr}
  \hline
Species & Model & MAE & 90\% Distance & Mean Obs. Cover \\ 
  \hline
BOGR & IPM & 12.18 & 38.52 & 9.43 \\ 
  BOGR & QBM & 19.66 & 56.50 & 9.26 \\ 
  HECO & IPM & 1.22 & 6.47 & 1.15 \\ 
  HECO & QBM & 12.35 & 41.11 & 1.18 \\ 
  PASM & IPM & 0.19 & 1.65 & 0.42 \\ 
  PASM & QBM & 0.55 & 7.78 & 0.42 \\ 
  POSE & IPM & 1.37 & 7.64 & 1.25 \\ 
  POSE & QBM & 1.79 & 40.59 & 1.27 \\ 
   \hline
\end{tabular}
\end{table}

@@

\begin{figure}[htbp]
\centering
\includegraphics{components/figure/manuscript-figure_2.pdf}
\caption{Proportional change in species' mean cover caused by a 1\%
increase in observed precipitation (+ppt), temperature (+temp), or both
(+ppt\&temp) as predicted by the individual-based IPM and the
aggregate-based QBM. Top panels show the mean predicted proportional
change in cover; lower panels show the 90\% quantiles of predicted
proportional cover change. Mean predictions and their associated
uncertainties are shown in separate panels for visual clarity.}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics{components/figure/manuscript-figure_3.pdf}
\caption{Sensitivity of equilibrium cover to a 1\% increase in
precipitation (+ppt), temperature (+temp), or both (+ppt\&temp) applied
to individual and combined vital rates. For example, the points
associated with G show the median cover from IPM simulations where a
climate perturbation is applied only to the growth regression climate
covariates.}
\end{figure}

@@

Clark, J. S., D. M. Bell, M. Kwit, A. Stine, B. Vierra, and K. Zhu.
2012. Individual-scale inference to anticipate climate-change
vulnerability of biodiversity. Philosophical Transactions of the Royal
Society B: Biological Sciences 367:236--246.
